---
title: "berms_analysis"
author: "Curtis R. Morris"
date: "12/03/25"
format:
  html:
    code-fold: true
    code-summary: "Code"
---

### 0. Set-up
```{r}
## Setup/Loading Data


## load libraries
pacman::p_load(dplyr,
               tidyverse,
               readr,
               glmmTMB, # zero inflation!
               car,
               broom.mixed,
               emmeans,
               vegan,
               easystats,
               knitr,
               kableExtra,
               broom,
               lubridate,
               performance
               )
## Helpers
theme_set(theme_classic(base_size = 14)+
            theme(axis.text.x = 
                    element_text(angle = 40, 
                                 vjust = 1, 
                                 hjust=1),
                  legend.position = "bottom"))

berm_control_colors <- c("#af8dc3", "#7fbf7b")

fix_data_for_plots <- function(dat){
  dat |>
    # make duxbury not have a _
    mutate(site = gsub("_", " ", site)) |>
    
    # make zone a factor and order so it's low, mid, high
    mutate( height = factor(height, levels = c("Low", "Mid", "High")))
}


## load data, bind them together, and add taxonomic data
quad_s23_dat <- read_csv("../data/quads_s23.csv", show_col_types = FALSE)
quad_f23_dat <- read_csv("../data/quads_f23.csv", show_col_types = FALSE)
quad_s24_dat <- read_csv("../data/quads_s24.csv", show_col_types = FALSE)
quad_s25_dat <- read_csv("../data/quads_s25.csv", show_col_types = FALSE)
seine_dat <- read_csv("../data/seine.csv", show_col_types = FALSE)
trap_dat <- read_csv("../data/traps.csv", show_col_types = FALSE)
farm_dat <- read_csv("../data/farms.csv", show_col_types = FALSE)
briv_dat <- read_csv("../data/brivs.csv", show_col_types = FALSE)
msl <- read_csv("../data/master_sp_list.csv", show_col_types = FALSE)

# Drop uneeded cols in msl and rename col
msl <- select(msl, 1:10) |> 
  rename(species_code = code)


```


### 1. Site-level Richness
```{r}
# bind data together, and drop the two NAs in Bayside Berm
richness_dat <- rbind(quad_s23_dat |> 
                        group_by(site, treatment) |> 
                        reframe(species_code = unique(species_code)),
                      quad_f23_dat |> 
                        group_by(site, treatment) |> 
                        reframe(species_code = unique(species_code)), 
                      quad_s24_dat |> 
                        group_by(site, treatment) |> 
                        reframe(species_code = unique(species_code)), 
                      quad_s25_dat |> 
                        group_by(site, treatment) |> 
                        reframe(species_code = unique(species_code)), 
                      seine_dat |> 
                        group_by(site, treatment) |> 
                        reframe(species_code = unique(species_code)),
                      trap_dat |> 
                        group_by(site, treatment) |> 
                        reframe(species_code = unique(species_code)),
                      briv_dat |> 
                        group_by(site, treatment) |> 
                        reframe(species_code = unique(first_predator)),
                      briv_dat |> 
                        group_by(site, treatment) |> 
                        reframe(species_code = unique(bait_consumer))
) |> 
  filter(species_code != is.na(species_code)) |> 
  filter(species_code != "NOSP") |> 
  filter(species_code != "none_present") |> 
  filter(species_code != "UCRA")

# attach taxonomic data
richness_dat <- left_join(richness_dat, 
                          msl, 
                          "species_code")

# Remove duplicate rows
richness_dat <- unique(richness_dat)

# Filter our codes to drop
richness_dat <- richness_dat |> 
  filter(
    !(site == "Bayside" & treatment == "Berm" & species_code %in% c("UFAC", "FUCU", "UBBL", "URAL")) &
    !(site == "Bayside" & treatment == "Control" & species_code %in% c("UFAC", "UFIS", "FUCU", "UBBL", "URFI", "URBL", "UCCA")) &
    !(site == "Coughlin" & treatment == "Berm" & species_code %in% c("UNAL", "UFAC", "FUCU", "UBAL", "UBBL", "UBFI", "URCB", "UCL", "URFC", "UCCA", "UBFC")) &
    !(site == "Coughlin" & treatment == "Control" & species_code %in% c("UNAL", "UFAC", "PHSP", "UCCA")) &
    !(site == "Duxbury_Beach" & treatment == "Berm" & species_code %in% c("UFAC", "PASP", "URFC")) &
    !(site == "Duxbury_Beach" & treatment == "Control" & species_code %in% c("UFAC", "URFC"))
  )


# Identify what species are found on berms and not on controls (and v.v.)
lostspp_dat1 <- richness_dat |> 
  filter(treatment == "Berm")

lostspp_dat2 <- richness_dat |> 
  filter(treatment == "Control")

# Species in lostspp_dat1 but not in lostspp_dat2
lost1 <- lostspp_dat1 |>
  anti_join(lostspp_dat2, by = c("site", "species_code"))

# Species in lostspp_dat2 but not in lostspp_dat1
lost2 <- lostspp_dat2 |>
  anti_join(lostspp_dat1, by = c("site", "species_code"))

# Combine both and remove temp df
lostspp_dat <- bind_rows(lost1, lost2)
rm(lostspp_dat1, lostspp_dat2, lost1, lost2)

# Calc richness per site_treatment
richness_dat <- richness_dat |>
  group_by(site, treatment) |>
  reframe(richness = n())

# Pivot for table
ttest_dat <- richness_dat |> 
  pivot_wider(names_from = "treatment",
              values_from = "richness") |> 
  mutate(delta = Berm-Control)


# Visualize
ggplot(richness_dat, 
                        mapping = 
         aes(x = treatment, 
             y = richness, 
             group = site,
             fill = treatment,
             shape = site)) +
  geom_line(color = "black") +
  geom_point(size = 3, 
             color = "black", 
             stroke = 1.2) +
  labs(title = "Site Level Species Richness by Treatment",
       x = "Treatment",
       y = "Species Richness") +
  scale_fill_manual(values = c("Berm" = "#af8dc3", "Control" = "#7fbf7b")) + 
  scale_shape_manual(values = c(21, 22, 24)) +
  guides(fill = "none")
```


```{r}
# Table of site level richness
ttest_dat |> 
  kable("html", caption = "Site Level Species Richness by Treatment") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")
```

Site-level richness at all sites and across all sampling methods and timepoints is higher in the berm treatment than in the control. Due to low sample size (n=3) these results are not statistically significant.

```{r}
# Table of species lost per site-treatments
lostspp_dat |> 
  filter(treatment == "Berm") |> 
  select(c(site, species_code, classification_description)) |> 
  kable("html", caption = "Species found only on the Berm Treatment") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")

lostspp_dat |> 
  filter(treatment == "Control") |> 
  select(c(site, species_code, classification_description)) |> 
  kable("html", caption = "Species found only on the Control Treatment") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")

```

Given results of later analysis and the nature of the specific species (highly mobile or rare, more tied to stochasticity than to an actual effect) that are missing from a giving treatment, I would argue that the story here is richness is unchanged between the treatments, i.e. no negative effect of berms (as opposed to greater richness on the berms). This is good, given the known decrease in richness/diversity/function of the biological communities found on grey infrastructure.

Also, the site with the greatest difference (duxbury) is wide and sandy. So we're comparing communities found on a rocky cobble shore (berm) to that on a dynamic sandy ecosystem. Our protocol does not sample interstitial spp, so richness is likely underestimated on Dux control.


### 2. Biodiversity, Cover Composition, & Functional Groups

```{r}
# Bind together
quad_dat <- rbind(quad_s23_dat,
                  quad_f23_dat,
                  quad_s24_dat,
                  quad_s25_dat)

# rm/ spaces in col titles and change none_present to NOSP, and change measurement type = NA to PErcent, also drop the duplicate entry of a quad
quad_dat <- quad_dat |> 
  mutate(measurement_type = if_else(measurement_type == "Percent Cover", "Percent", measurement_type))

quad_dat <- quad_dat |> 
  mutate(species_code = if_else(species_code == "none_present", "NOSP", species_code),
         measurement_type = if_else(species_code == "NOSP" & is.na(measurement_type),"Percent", measurement_type),
         height = factor(height, levels = c("Low", "Mid", "High"))) |> 
  slice(-808)


# Drop data that are no longer needed
rm(quad_s23_dat)
rm(quad_f23_dat)
rm(quad_s24_dat)
rm(quad_s25_dat)

# Store total quad count for later use
n_total <- quad_dat |>
group_by(site, treatment, season, height) |>
  summarise(quad_count = n_distinct(quadrat), .groups = "drop") |>
  summarise(n_total = sum(quad_count))

# Make sure every quad with 0 has at a NOSP place holder for percent and for count
quad_dat <- quad_dat |> 
  filter(species_code == "NOSP") |> 
  group_by(season, site, treatment, height, quadrat) |> 
  complete(measurement_type = c("Percent", "Count"),
    fill = list(value = 0, species_code = "NOSP")) |> 
  ungroup() |>
  bind_rows(quad_dat %>% filter(species_code != "NOSP"))

# quad_dat |>
#   filter(species_code == "NOSP") |>
#   count(season, site, treatment, height, quadrat) |> 
#   View()
# All there!

# Sum to the quad level
quad_dat <- quad_dat |> 
  group_by(season, site, treatment, height, quadrat, measurement_type, species_code) |> 
  reframe(sum = sum(measurement)) |> 
  ungroup()

# Pivot data for use in vegan::diversity() 
quad_dat <- quad_dat |> 
  pivot_wider(names_from = "species_code",
              values_from = "sum",
              values_fill = 0)

# Drop placeholder spp for quads w/ nothing 
quad_dat <- quad_dat |> 
  select(-c("NOSP"))

##
#  Richness and Shannon Biodiversity analysis ----------------------------------
##

# Calculate Shannon, InvSimpson, richness, evenness, and rearrange
div_dat <- quad_dat |> 
  mutate(shannon = diversity(quad_dat[,-c(1:6)], index = "shannon")) # measures the uncertainty in predicting the species identity of an individual chosen at random from the community. A higher value indicates higher diversity (more species and/or more even distribution), and a lower value indicates lower diversity. It equals 0 only when there is no uncertainty—meaning there is only ONE species present.

div_dat <- div_dat |>  # not going to analyze. Shannon and richness should suffice
  mutate(invsimpson = diversity(div_dat[,-c(1:6,85)], index = "invsimpson")) 

div_dat <- div_dat |> 
  mutate(richness = specnumber(div_dat[,-c(1:6,85,86)])) 

div_dat <- div_dat |>  # Going to ignore this in anlysis since they're similar and I think this will be better captured by the analysis in the cover composition section
  mutate(evenness = shannon/log(richness)) # Pielous eveness how evenly individuals are distributed across the different species in a community. It ranges from 0 (completely uneven) to 1 (perfectly even).# Evenness of 0 occurs when richness is 1 (i.e. log 1 = 0, dividing by 0 = NaN)
  
div_dat <- div_dat |> 
select(richness, shannon, evenness, invsimpson, everything())
  
# Set up data for plotting
div_plot_dat <- div_dat[,-c(11:88)] |> 
  pivot_longer(cols = richness:invsimpson, 
               values_to = "value", 
               names_to = "index")

div_plot_dat$height <- factor(div_plot_dat$height,
                              levels = c("Low", "Mid", "High"))


##
# Analysis
##

## Shannon
# Hurdle Model - Odds of being over 0
shannon_percent_hurdle_mod <- glm(shannon > 0 ~ treatment + height + site + treatment:height,
                                  family = binomial(link = "logit"),
                                  data = div_dat |> filter(measurement_type == "Percent"))

shannon_count_hurdle_mod <- glm(shannon > 0 ~ treatment + height + site + treatment:height,
                                  family = binomial(link = "logit"),
                                  data = div_dat |> filter(measurement_type == "Count"))

# Models for those over 0
shannon_percent_posthurdle_mod <- lm(sqrt(shannon) ~ treatment + height + site + treatment:height, # note that shannon at richness of 1 is 0, no diversity, only 1 sp present which is the same as richness of 0
                          data = div_dat |> filter(measurement_type == "Percent", shannon > 0))

shannon_count_posthurdle_mod <- lm(sqrt(shannon) ~ treatment + height + site + treatment:height,
                        data = div_dat |> filter(measurement_type == "Count", shannon > 0))

# Get model stats
# check_model(shannon_percent_hurdle_mod) # Homogeneity of variance seems to be the only problem
# check_model(shannon_count_hurdle_mod) # Homogeneity of variance, also high collinearity for interaction parameter
# check_model(shannon_percent_posthurdle_mod) # PPC could be better, Homogeneity of variance not great
# check_model(shannon_count_posthurdle_mod) # treatment and treatment:height has high colineaerity

## So what does this tell us
# Hurdle (0,1) for shannon percent
# Anova(shannon_percent_hurdle_mod) |> tidy() # height plys biggest roll, but site and treatment also play a roll
# tidy(shannon_percent_hurdle_mod)
# 
# shannon_percent_hurdle_em <- emmeans(shannon_percent_hurdle_mod,
#                     ~treatment | height)
# 
# contrast(shannon_percent_hurdle_em, method = "pairwise") # Significant effect of treatment in the low and mid


# Hurdle (0,1) for shannon count
# Anova(shannon_count_hurdle_mod) |> tidy() # all important, by again height the most so
# tidy(shannon_count_hurdle_mod) # need to back transform the estimate?
# 
# shannon_count_hurdle_em <- emmeans(shannon_count_hurdle_mod,
#                                      ~treatment | height)
# 
# contrast(shannon_percent_hurdle_em, method = "pairwise") # Significant effect of treatment in the low and mid

# Post-hurdle (>0) for shannon percent
# Anova(shannon_percent_posthurdle_mod) |> tidy() # Percent Shannon  differs significantly across heights and sites, but is not strongly affected by treatment, and the interaction between treatment and height is also not significant.
# tidy(shannon_percent_posthurdle_mod) |>  
#   mutate(estimate_original = sign(estimate) * estimate^2,
#          std.error_original = 2 * abs(estimate) * std.error) |> 
#   mutate(lower.CL = ((estimate_original - std.error_original)^2),
#          upper.CL = ((estimate_original + std.error_original)^2))

# Height matters a lot: Low  plots have the highest Shannon diversity (0.768 and 0.535).Mid heights are intermediate (0.511 and 0.375). High height plots have the lowest diversity (0.070 and 0.049).
# Treatment effect is small: Within each height category, Berm has slightly higher Shannon diversity than Control.
# So these differences are statistically robust (all p values are << .001.)
# Differences between heights are much larger than differences between treatments. Matches ANOVA: height is the main driver, treatment is minor.
# 
# shannon_percent_posthurdle_em <- emmeans(shannon_percent_posthurdle_mod,
#                                    ~treatment | height)
# 
# contrast(shannon_percent_posthurdle_em, method = "pairwise") # low and mid treatment has a significant effect

# Post-hurdle (>0) for shannon count
# Anova(shannon_count_posthurdle_mod) |> tidy() # essentially the same as percent
# tidy(shannon_count_posthurdle_mod) |>  # height has largest effect, while there is some decrease between treatment
#   mutate(estimate_original = sign(estimate) * estimate^2, # note that the negative estimates are lost after the back transformation, so i've multiplioed them by the sign to return the original sign
#          std.error_original = 2 * abs(estimate) * std.error) |> 
#   mutate(lower.CL = ((estimate_original - std.error_original)^2),
#          upper.CL = ((estimate_original + std.error_original)^2))

# 
# shannon_count_posthurdle_em <- emmeans(shannon_count_posthurdle_mod,
#                                    ~treatment | height)
# 
# contrast(shannon_count_posthurdle_em, method = "pairwise") # no significance

## richness
# Hurdle Model - Odds of being over 0
richness_percent_hurdle_mod <- glm(richness > 0 ~ treatment + height + site + treatment:height,
                                   family = binomial(link = "logit"),
                                   data = div_dat |> filter(measurement_type == "Percent"))

richness_count_hurdle_mod <- glm(richness > 0 ~ treatment + height + site + treatment:height,
                                 family = binomial(link = "logit"),
                                 data = div_dat |> filter(measurement_type == "Count"))

# Models for those over 0
richness_percent_posthurdle_mod <- lm(sqrt(richness) ~ treatment + height + site + treatment:height,
                                      data = div_dat |> filter(measurement_type == "Percent", richness > 0))

richness_count_posthurdle_mod <- lm(sqrt(richness) ~ treatment + height + site + treatment:height,
                                    data = div_dat |> filter(measurement_type == "Count", richness > 0))

# Get model stats
# check_model(richness_percent_hurdle_mod)
# check_model(richness_count_hurdle_mod)
# check_model(richness_percent_posthurdle_mod)
# check_model(richness_count_posthurdle_mod) 
# 
# ## So what does this tell us
# # Hurdle (0,1) for richness percent
# Anova(richness_percent_hurdle_mod) |> tidy() # all sig, height > treatment > site
# tidy(richness_percent_hurdle_mod)
# 
# richness_percent_hurdle_em <- emmeans(richness_percent_hurdle_mod,
#                                      ~treatment | height)
# 
# contrast(richness_percent_hurdle_em, method = "pairwise") # Significant effect of treatment in all (results identical?)
# 
# 
# # Hurdle (0,1) for richness count
# Anova(richness_count_hurdle_mod) |> tidy() # all sig, height > treatment > site
# tidy(richness_count_hurdle_mod) # effect of treatment and height evident
# 
# richness_count_hurdle_em <- emmeans(richness_count_hurdle_mod,
#                                    ~treatment | height)
# 
# contrast(richness_count_hurdle_em, method = "pairwise") #Significant effect of treatment in all (results identical?)
# 
# # Post-hurdle (>0) for richness percent
# Anova(richness_percent_posthurdle_mod) |> tidy() # effect of height and site, not treatment 
# tidy(richness_percent_posthurdle_mod) |>  
#   mutate(estimate_original = sign(estimate) * estimate^2,
#          std.error_original = 2 * abs(estimate) * std.error)|> 
#   mutate(lower.CL = ((estimate_original - std.error_original)^2),
#          upper.CL = ((estimate_original + std.error_original)^2))
# 
# richness_percent_posthurdle_em <- emmeans(richness_percent_posthurdle_mod,
#                                          ~treatment | height)
# 
# contrast(richness_percent_posthurdle_em, method = "pairwise") #no sig
# 
# # Post-hurdle (>0) for richness count
# Anova(richness_count_posthurdle_mod) |> tidy() # height and site
# 
# tidy(richness_count_posthurdle_mod) |>  # low zone and duxbury are particularlly impactful
#   mutate(estimate_original = sign(estimate) * estimate^2, # note that the negative estimates are lost after the back transformation, so i've multiplioed them by the sign to return the original sign
#          std.error_original = 2 * abs(estimate) * std.error) |> 
#   mutate(lower.CL = ((estimate_original - std.error_original)^2),
#          upper.CL = ((estimate_original + std.error_original)^2))
# 
# 
# richness_count_posthurdle_em <- emmeans(richness_count_posthurdle_mod,
#                                        ~treatment + height)
# 
# contrast(richness_count_posthurdle_em, method = "pairwise") # no significance
# 


# Shannon, post-hurdle
## by percent cover

# Summarize mean and standard error for plotting
plot_data <- div_dat %>%
  filter(measurement_type == "Percent") %>%
  group_by(height, site, treatment) %>%
  summarise(
    mean_shannon = mean(shannon, na.rm = TRUE),
    se_shannon = sd(shannon, na.rm = TRUE)/sqrt(n()),
    .groups = "drop") |> 
  mutate(height = factor(height, levels = c("Low", "Mid", "High")))

# Plot that shows shannon decreasing by zone and site, but relatively unchanged by treatment
ggplot(plot_data, aes(x = height, y = mean_shannon, fill = site)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = mean_shannon - se_shannon, ymax = mean_shannon + se_shannon),
                width = 0.2, position = position_dodge(width = 0.8)) +
  facet_wrap(~treatment) +
  labs(
    y = "Shannon Diversity",
    x = "Height",
    title = "Effect of Height and Site on Shannon Diversity for Percent Cover")
```



```{r}
# Plot that shows shannon unchanged by treatment
ggplot(plot_data, aes(x = height, y = mean_shannon, fill = treatment)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = mean_shannon - se_shannon, ymax = mean_shannon + se_shannon),
                width = 0.2, position = position_dodge(width = 0.8)) +
  facet_wrap(~site) +
  labs(
    y = "Shannon Diversity",
    x = "Height",
    title = "Effect of Treatment on Shannon Diversity for Percent Cover")

## by count

# Summarize mean and standard error for plotting
plot_data <- div_dat %>%
  filter(measurement_type == "Count") %>%
  group_by(height, site, treatment) %>%
  summarise(
    mean_shannon = mean(shannon, na.rm = TRUE),
    se_shannon = sd(shannon, na.rm = TRUE)/sqrt(n()),
    .groups = "drop") |> 
  mutate(height = factor(height, levels = c("Low", "Mid", "High")))

# Plot that shows shannon decreasing by zone and site, but relatively unchanged by treatment
ggplot(plot_data, aes(x = height, y = mean_shannon, fill = site)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = mean_shannon - se_shannon, ymax = mean_shannon + se_shannon),
                width = 0.2, position = position_dodge(width = 0.8)) +
  facet_wrap(~treatment) +
  labs(
    y = "Shannon Diversity",
    x = "Height",
    title = "Effect of Height and Site on Shannon Diversity for Count")

# Plot that shows shannon unchanged by treatment
ggplot(plot_data, aes(x = height, y = mean_shannon, fill = treatment)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = mean_shannon - se_shannon, ymax = mean_shannon + se_shannon),
                width = 0.2, position = position_dodge(width = 0.8)) +
  facet_wrap(~site) +
  labs(
    y = "Shannon Diversity",
    x = "Height",
    title = "Effect of Treatment on Shannon Diversity for Count")
```

For presence/absence of any species for both percent and count data, height, site, and treatment all had a significant effect on shannon diversity, and height explained the largest amount of variance. When averaged across all three sites, the effect of treatment was significant only in the low and mid zones.

When species were present for both percent and count data, shannon  differs significantly across both height and site, but is not strongly affected by treatment, and the interaction between treatment and height is also not significant. Height matters a lot: Low  plots have the highest Shannon diversity (0.768 and 0.535).Mid heights are intermediate (0.511 and 0.375). High height plots have the lowest diversity (0.070 and 0.049). Treatment effect is small: Within each height category, Berm has slightly higher Shannon diversity than Control and is non-significant.

```{r}
# Richness, post-hurdle
## by percent cover

# Summarize mean and standard error for plotting
plot_data <- div_dat %>%
  filter(measurement_type == "Percent") %>%
  group_by(height, site, treatment) %>%
  summarise(
    mean_richness = mean(richness, na.rm = TRUE),
    se_richness = sd(richness, na.rm = TRUE)/sqrt(n()),
    .groups = "drop") |> 
  mutate(height = factor(height, levels = c("Low", "Mid", "High")))

# Plot that shows richness decreasing by zone and site, but relatively unchanged by treatment
ggplot(plot_data, aes(x = height, y = mean_richness, fill = site)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = mean_richness - se_richness, ymax = mean_richness + se_richness),
                width = 0.2, position = position_dodge(width = 0.8)) +
  facet_wrap(~treatment) +
  labs(
    y = "Species Richness",
    x = "Height",
    title = "Effect of Height and Site on Species Richness for Percent Cover")

# Plot that shows richness unchanged by treatment
ggplot(plot_data, aes(x = height, y = mean_richness, fill = treatment)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = mean_richness - se_richness, ymax = mean_richness + se_richness),
                width = 0.2, position = position_dodge(width = 0.8)) +
  facet_wrap(~site) +
  labs(
    y = "Species Richness",
    x = "Height",
    title = "Effect of Treatment on Species Richness for Percent Cover")

## by count

# Summarize mean and standard error for plotting
plot_data <- div_dat %>%
  filter(measurement_type == "Count") %>%
  group_by(height, site, treatment) %>%
  summarise(
    mean_richness = mean(richness, na.rm = TRUE),
    se_richness = sd(richness, na.rm = TRUE)/sqrt(n()),
    .groups = "drop") |> 
  mutate(height = factor(height, levels = c("Low", "Mid", "High")))

# Plot that shows richness decreasing by zone and site, but relatively unchanged by treatment
ggplot(plot_data, aes(x = height, y = mean_richness, fill = site)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = mean_richness - se_richness, ymax = mean_richness + se_richness),
                width = 0.2, position = position_dodge(width = 0.8)) +
  facet_wrap(~treatment) +
  labs(
    y = "Species Richness",
    x = "Height",
    title = "Effect of Height and Site on Species Richness for Count")

# Plot that shows richness unchanged by treatment
ggplot(plot_data, aes(x = height, y = mean_richness, fill = treatment)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(aes(ymin = mean_richness - se_richness, ymax = mean_richness + se_richness),
                width = 0.2, position = position_dodge(width = 0.8)) +
  facet_wrap(~site) +
  labs(
    y = "Species Richness",
    x = "Height",
    title = "Effect of Treatment on Species Richness for Count")
```

For percent species richness hurdle (i.e. presence absence), there was a significant effect of all of height, treatment, site, and the treatment:height interaction term. The results for count were similar, save that the interaction term was non-significant.

Post-hurdle, height and site, but not treatment, played a significant role in determining richness.

### 5. Herbivore Abundance (LILI)
```{r}
## load the data and filter to Littorina littorea
## then sum for each quad
snail_dat <- read_csv("../data/quads_s23.csv", show_col_types = FALSE) |>
  filter(measurement_type == "Count") |>
  filter(species_code == "LILI") |>
  group_by(site, treatment, quadrat, height) |>
  summarize(snails = sum(measurement, na.rm = TRUE)) |>
  ungroup() |> fix_data_for_plots()

## Fix up snail data to fill in 0 for quads where there
## were no snails
snail_dat <- snail_dat |>
  complete(site, treatment, nesting(quadrat, height),
           fill = list(snails = 0)) |>
  mutate(are_there_snails = as.numeric(snails !=0))


## an initial plot
ggplot(snail_dat,
       aes(x = site, color = treatment,
           y = snails)) +
  geom_point(position = position_dodge(width = 0.5),
             alpha = 0.5,
             size = 2) +
  scale_color_manual(values = berm_control_colors) +
  facet_wrap(vars(height)) +
  labs(x="", 
       y = "Littorina per\n0.25 sq m",
       color = "") 


# what's the distribution of snails?
ggplot(snail_dat,
       aes(x = snails, fill = site)) +
  geom_histogram() +
  facet_grid(rows = vars(treatment),
             cols = vars(height)) +
  labs(x = "Littorina per\n0.25 sq m")
```

Difficult to impossible to analyze statistically due to complete seperation - i.e. presence of site:treatments that are completely 0 (for example, no LILI in Bayside|Duxbury:Control for the High and Mid). But we can analyze for the low zone at Coughlin and Bayside which contain non-zero values at both treatments.

```{r}
##
# Analysis of where are there even snails
##
# snail_pres_mod <- glm(are_there_snails ~
#                         height * treatment*site,
#                       data = snail_dat,
#                       family = binomial)
# 
# Anova(snail_pres_mod) |> tidy() |>
#   mutate(p.value = round(p.value, digits = 10))

## Complete separation problem - write about things
## without stats in the text!

##
# Analysis of Low tide height
##
## an analysis using zero inflated negative binomial
## just for the low where we don't have site:treatments
## that are completely 0
snail_mod <- glmmTMB(snails ~ treatment*site,
                     zi = ~  site,
                     data = snail_dat |> filter(height == "Low"),
                     family = nbinom2)

# model diagnostics
# performance::check_model(snail_mod)


# summarizing tests
Anova(snail_mod) |> tidy()
tidy(snail_mod)

# posthoc
snail_em <- emmeans(snail_mod,
                    ~treatment | site)

snail_pairwise <- contrast(snail_em, method = "pairwise") |> 
  as.data.frame()

kable(snail_pairwise, digits = 3, caption = "LILI Abundance Contrasts by Site")
# Must ignore Dux results because there were no LILI in the low control treatment
```

Our model predicts an increase of 1.29 LILI when going from the Control to Berm in the low zone of Bayside. Duxbury cannot be determined statistically, however, this is  due to the absence of any LILI in the control. Results for Coughlin were not statistically significant.

These results align with what we see at the sites. At Bayside, the presence of the berm modifies the substrate such that the control is characterized as a small grain sand-shell hash ecosystem, and the berm has ample habitat as provisioned by the cobbles. The cobbles are also present at coughlin, however their ecological effect on provisioning habitat/ameliorating stress and thus increasing LILI abundance is decreased because the preexisting substrate is already large grain cobble in nature, so it's not modifying the preexisting substrate of the area and thus does not affect LILI abundance in a significant way. Duxbury cannot be analyzed due to the absence of LILI, but again, I would argue this is due to underlying differences in substrate among sites amd thus the impact of the berm. Duxbury is a sand ecosystem. The installation of the berm fundamentally changes this system, such that without the berm LILI simply are not present (instead Illynasa is, a soft sediment snail). 


### 6. Mobiles: Traps and Seine

In 2023 we assessed mobile species presence and abundance using baited crab traps, unbaited conical fish traps, and seine net tows conducted parallel to the shore.
```{r}
##
# Exploring the data
##

# Abundance of each spp per trap
abundance_dat <- trap_dat |>
    group_by(site, treatment, trap_name, trap_number, species_code) |>
    reframe(abundance = n()) |> 
  pivot_wider(names_from = species_code, values_from = abundance, values_fill = 0) |> 
  select(-NOSP)

# CAMA only
CAMA_dat <- abundance_dat |>
  select(1:5) |> 
  filter(trap_name == "Crab") |> 
  select(-trap_name)

# All spp for table
abundance_dat <- abundance_dat |>
  group_by(site, treatment, trap_name) |> 
  summarise(
    n = n_distinct(trap_number),
    across(CAMA:FUNS, ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  )

abundance_dat |> 
  kable("html", caption = "Species abundance by site, treatment, and trap type") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")
```

Across all sites and treatments Carcinus maenus was the most abundant catch. Hemigrapsus occurred as a single individual in three traps, and all other species occurred only as a single individual. We will analyze only the abundance of CAMA as a function of treatment.

```{r}

##
# Visualize
##

# CAMA abundance by site as a function of treatment
ggplot(CAMA_dat, 
                    mapping = aes(x = treatment, 
                              y = CAMA)) +
  geom_boxplot() +
  geom_point(position = "jitter",
              alpha = .6)+
  labs(title = "CAMA abundance in traps",
       x = "Treatment",
       y = "CAMA Abundance") +
    facet_wrap(vars(site))

##
# Stats
##

# ttest for CAMA abundnace
  CAMA_ttest <- t.test(CAMA ~ treatment, data = CAMA_dat, var.equal = FALSE) |> 
    tidy()

CAMA_ttest |> 
  kable("html", caption = "T-test results for CAMA abundance by treatment") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")
  
  
# estimate estimate1 estimate2 statistic p.value parameter conf.low   conf.high 
# 1.73      6.87      5.13     0.573   0.572      24.5    -4.51      7.97
#   
```
Conclusions: no differences in CAMA abundance in a trap given a treatment. 
 
Results will now be presented for the Seine tows. 8 tows were conducted at Bayside (4 per treatment), however only 2 were conducted at Coughlin and Duxbury (1 per treatment). Data will be presented for all tows, however due to sample size only the resutls from Bayside will be analyzed.
 
```{r}
# Add date column
seine_dat <- seine_dat |> 
  mutate(date = make_date(year, month, day))

# Estimate Volume sampled
seine_dat <- seine_dat|>
  group_by(site, treatment, date) |>
  mutate(vol_sampled = ((estimated_depth_pole_1_m/estimated_depth_pole_2_m)/2)*width_net_aperture_m*distance_walked_m)

# Drop cols not interested in anymore
seine_dat <- seine_dat |> 
  select(-c(3:17,19:22))

# Reframe for abundance per m3 sampled
seine_dat <- seine_dat|>
  group_by(site, treatment, date, vol_sampled, species_code) |>
  reframe(abundance = n()) |> 
  mutate(concentration = abundance/vol_sampled)

# Make a table showing concentration by species for each tow
seine_table_dat <- seine_dat |> 
  select(-abundance) |> 
  mutate(concentration_100m3 = concentration*100) |>
  select(-concentration) |> 
  pivot_wider(names_from = species_code, 
              values_from = concentration_100m3,
              values_fill = 0) |> 
  mutate(across(where(is.numeric), ~ ifelse(.x == 0, "0", format(round(.x, 2), nsmall = 2))))

seine_table_dat |> 
  kable("html", caption = "Seine Haul Species Abundance per 100 m3", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")
```
As with the traps, at all sites and treatments Carcinus maenus was the most abundant catch and thus is the only species to be included in the analysis below. 

```{r}
##
# Visualize
##

# Select only bayside data, other sites have no replication
seine_bayside_dat <- seine_table_dat |> 
  filter(site == "Bayside") |> 
  ungroup()

# Make numberic
seine_bayside_dat$CAMA <- as.numeric(seine_bayside_dat$CAMA)


# CAMA abundance at bayside as a function of treatment
#seine_plot <- 
  ggplot(seine_bayside_dat, 
                    mapping = aes(x = treatment, 
                                  y = CAMA)) +
  geom_boxplot() +
  labs(title = "CAMA per 100 m3",
       x = "Treatment",
       y = "Concentration (indiv/100m3)")

##
# Stats
##

# ttest for CAMA abundnace at bayside
CAMA_seine_dat <- seine_dat |> 
  filter(species_code == "CAMA")

CAMA_seine_ttest <- t.test(data = seine_bayside_dat, 
                     CAMA ~ treatment, var.equal = FALSE) |> 
  tidy()


CAMA_seine_ttest |> 
  kable("html", caption = "T-test results for CAMA abundance in Seine tows at Bayside") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")

# estimate estimate1 estimate2 statistic p.value parameter conf.low conf.high      
# -6.92      4.06      11.0     -1.39   0.221      5.15    -19.6      5.76 

# Not significant
```

More CAMA were captured in hauls done at the control site, however results were not signficant.


### 7. Predation: BRIVS

```{r}

## Verify same amount of brivs per site-treatment
# briv_dat |>
#   group_by(site, treatment) |>
#   reframe(briv_count = n()) |>
#   View()
# Not quite even, Coughlin has 5 in each, bayside is 9/8, dux is 10/10


# Clean up data
briv_scrub_dat <- briv_dat |> 
  select(52, everything ()) |> 
  filter(notes != "No visibility") |> 
  filter(notes != "video data not usable") |> 
  filter(notes != "No data, camera did not record") |> 
  filter(notes != "No Visibility") |> 
  filter(notes != "No visibility for first 2 videos, count started after bait was eaten") |>
  filter(notes != "No Visibility, cannot see bait for much of video, benthos cannot be seen at all. CAMA at 1098 was persistent and potentially consumer, however visibility prevents conclusive determination.") |>
  filter(notes != "No Visibility, cannot see bait or benthos") |>
  filter(notes != "No Visibility, cannot see bait or benthos; bait quickly targeted by CAMA, which dissapears and reappears frequently due to poor visibility. Likely only one individual, and this islikely the consumer of bait. Impossible to confirm the moment that the bait is actualy removed entirely.") |> 
  filter(notes != "No visibility. A crab appears at somepoint clasped to the BRIV and feeding on squid. Unknown when it first appears, whether it eats the bait completly, and what it is.") |> 
  filter(notes != "No visibility for much of video.Unknown what ultimatly consumes the bait. Likely CAMA but cannot be confirmed from the video")
  
# briv_scrub_dat |> 
#   group_by(site, treatment) |> 
#   summarise(n = n())


briv_dat |> 
  group_by(site, treatment, first_predator) |>
  filter(!is.na(first_predator)) |> 
  filter(first_predator != "NOSP") |> 
  summarise(n = n())
  
briv_dat |> 
  group_by(site, treatment, bait_consumer) |>
  filter(!is.na(bait_consumer)) |> 
  filter(bait_consumer != "NOSP") |> 
  summarise(n = n())

```
When scrubbed of poor quality video, no replication remains for predation analysis. 

For usable videos, save for two instances where an unknown crab could not be identified, the first predator was always CAMA, across all sites and treatments. Likewise, bait consumer was always CAMA. 



### 8. Crypto-benthic Fish: FARMS
```{r}
# (re)load data
farm_dat <- read_csv("../data/farms.csv")

# Work it up
farm_dat <- farm_dat |> 
  mutate(date_deployed = make_date(year_deployed, month_deployed, day_deployed)) |>
  mutate(date_retrieved = make_date(year_retrieved, month_retrieved, day_retrieved)) |>
  mutate(soak_duration = date_retrieved - date_deployed) |> 
  select(-c(4:14,16:20)) |> 
  select(1,2,3,5,6,7,4) |> 
  group_by(site, treatment, farms_number, date_deployed, date_retrieved, soak_duration, species_code) |>
  reframe(abundance = n()) |> 
  pivot_wider(names_from = species_code, values_from = abundance, values_fill = 0) |> 
  filter(!is.na(date_retrieved)) |> 
  select(-c(NOSP, LESP, CRFO, ULSL, LILI)) |> # get rid of nothing placeholder, ULSL is supposed to be ULSP which is an algae, not sure why it was recorded, CRFO only recoreded on one (i.e. not consistently recorded), ditto for lesp
  mutate(soak_class = if_else(condition = soak_duration > 100, "long", "short")) |> 
  rename(FUNS = FUSP) # correcting this code. FUSP is an algae, FUNS is a fish


# Calc abundance found per farm by retrieval season/soak duration
farm_soak_dat <- farm_dat |>
  select(1,2,19, 7:18) 

farm_soak_dat <- farm_soak_dat |> 
  group_by(site, treatment, soak_class) |> 
  summarise(
    n_farms = n(),
    across(
      where(is.numeric),
      \(x) sum(x, na.rm = TRUE)), # sum up across farms
    .groups = "drop"
  ) |>
  mutate(
    across(
      -c(1:4),
      ~ round(.x / n_farms,2))) # standardize by count per farm

# Table of FARM occupancy
farm_soak_dat |> 
  kable("html", caption = "FARM occupancy by soak duration") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")
```
 FPEN, CAMA, PHGU, PAHE are the codes that occur in abundance
 FPEN most abundant in the long soak, but still present in the short soak
 PAHE only in winter/long soak, and PHGU mostly occurs in winter/long soak. This could be a factor of soak duration or seasonal displacement/migration of species.

 For this analysis, I will focus only on those spp which occur at some abundance (FPEN, CAMA, PHGU, PAHE)
 Also excluding coughlin due to not enough farms recovered
 
 # CAMA

```{r}
# CAMA
CAMA_mod <- glmmTMB(CAMA ~ treatment + soak_class, 
                     data = farm_dat |> filter(site == "Bayside"), 
                     family = poisson(link = "log"))

# check_model(CAMA_mod)
# 
# summarizing tests
Anova(CAMA_mod) |> tidy() # Soak class seems to explain more than the treatment
tidy(CAMA_mod)

ggplot(farm_dat |> filter(site == "Bayside"), 
                    mapping = aes(x = treatment, 
                                  y = CAMA,
                                  fill = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(shape = soak_class),  
              width = 0.2,
              height = 0, 
              alpha = 0.6, 
              size = 2) +
  labs(title = "CAMA abundance in farms",
       x = "Treatment",
       y = "CAMA Abundance") +
  facet_wrap(vars(soak_class))
```

Soak duration explains more of CAMA abundance than treatment, however neither had a statistically significant effect.

# FPEN

```{r}
# FPEN
FPEN_mod <- glmmTMB(FPEN ~ treatment + soak_class, 
                    data = farm_dat |> filter(site == "Bayside"), 
                    family = poisson(link = "log"))

# check_model(FPEN_mod)
# 
# # summarizing tests
Anova(FPEN_mod) |> tidy()
tidy(FPEN_mod)
# Fewer shrimps in the control and in the short soak time. Excluded coughlin due to low soak time

ggplot(farm_dat |> filter(site == "Bayside"), 
       mapping = aes(x = treatment, 
                     y = FPEN,
                     fill = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(shape = soak_class),  
              width = 0.2,
              height = 0, 
              alpha = 0.6, 
              size = 2) +
  labs(title = "FPEN abundance in farms",
       x = "Treatment",
       y = "FPEN Abundance") +
  facet_wrap(vars(soak_class))
```
Fewer shrimps in the short soak time/winter time point. Fewer shrimp in the control treatment, though this is not reflected in the winter timepoint.

# PHGU

```{r}
# PHGU - Bayside only
PHGU_mod <- glmmTMB(PHGU ~ treatment + soak_class, 
                    data = farm_dat |> filter(site == "Bayside"), 
                    family = poisson(link = "log"))



# check_model(PHGU_mod)
# 
# # summarizing tests
Anova(PHGU_mod) |> tidy()
tidy(PHGU_mod)
# Lose PHGU in the control and the short


ggplot(farm_dat |> filter(site == "Bayside"), 
       mapping = aes(x = treatment, 
                     y = PHGU,
                     fill = treatment)) +
  geom_boxplot() +
  geom_jitter(aes(shape = soak_class),  
              width = 0.2,
              height = 0, 
              alpha = 0.6, 
              size = 2) +
  labs(title = "PHGU abundance in farms",
       x = "Treatment",
       y = "PHGU Abundance") +
  facet_wrap(vars(soak_class))

```
Both treatment and soak class affected PHGU abundance, with the short class and the control haveing fewer individuals.  
